Eye Fitting Straight Lines in the Modern Era

### 2022-08-03 manuscript-revisions/response-to-reviewers.qmd

  + \response{Capitalized all "grammar of graphics" to "Grammar of Graphics".}
+ p.6, l.42: use single quotes for `You Draw It` as in the abstract
  + \response{Added single quotes to 'You Draw It'.}
+ p.7, top: be specific and indicate in which of the sections of your paper each of these points is being addressed, in other words, provide a specific overview what can be found in Sections 2 - 5 of your paper (and also mention the supplementary materials in 1 or 2 sentences)
  + \response{Added information about specific parts contained in each of the paper sections.}
+ p.7, l.34: you had 35 participants and got 119 plots; later (p.10, l.38) you state that each participant had to evaluate 4 plots; provide this information here immediately; can you summarize why there are 21 plots missing, e.g., some plots were incomplete, some participants were dropping out after just 1 or 2 plots, etc.
  + \response{After double checking the data, it was actually 131 plots - we are working with pilot data and thus had this number from another set. We clarified and made this number now reproducible by using r-code in the rmd template file rather than mannually entering it. We also discussed the missing plots by adding: "Data were collected as a part of a pilot study meant to test the applet; therefore, either voluntary participant dropout or disconnection from a server not designed to accommodate large magnitudes of participants resulted in missing plots in our data set for analysis."}

@@

+ p.8, l.55: "i = 1, ...N" -> "i = 1, ..., N,"
  + \response{Added "," into indexing on p11, l34 and p7, l55}
+ \dubcheck{p.9, top: $y_{\bar{x}}$ in Table 1 and $\beta_0$ in (1) are not properly defined and used; are they the same?}
  + \response{We aimed to replicate the trends from Mosteller et al. (1981). This original manuscript gives vague descriptions of the coefficients used to describe their trends. The way they defined their parameters was by using the y-value at the mean of the x-values $(y_{\bar{x}})$ and the slope. When simulated our data, we first generate and jitter x-values across the domain, therefore, x-bar may differ for each unique data set due to the random jitter. In turn, the actual y-intercept $(\beta_0)$ will vary slightly from data set to data set, but each trend will go through the same $(\bar{x}, y_{\bar{x}})$ point. We changed the equation notation to follow the point-slope equation rather than changing notation.}
+ p.9, l.25-37: can you further describe these four models with some more details and extend the table, e.g., list the domains in the table as well (and not only in the text); also indicate how many points were used in each scatterplot (e.g., was this random or fixed?; plot V in Fig 2 seems to have much fewer points than the other three); do the 4 letters have a certain meaning?
  + \response{We double checked the data simulation process based on the shiny applet code. There was an extra line of code in the manuscript file that incorrrectly filtered points out. There are were 30 points simulated for each parameter choice which is stated on p7 l53. A new "set" of data was simulated randomly for each participant with the same underlying parameters and a fixed number of points. We added the domain for each parameter choice to the table. The parameter choice names were selected to reflect the labels used in Mosteller et al. (1981). We clarified this on p8 l25 with "Model equation parameters, $beta_1$, $y_{\bar{x}}$, and parameter choice letter names (F, S, V, N) and V), were selected to reflect the four data sets used and labeled in Mosteller et al..." See analysis in supplementary material for code on the data simulation procedure.}

@@

+ p.10, l.38-39: I suppose each participant got one plot of type S, F, V, N ? - say so
  + \response{We clarified that each participant got one of each of the plot types in Section 2.4.}

@@

+ p.11, l.35 & l.48: be consistent, e.g., "increment" vs. "increments" and missing specification of k
in the 2nd part
  + \response{Changed to "increment" for both to be consistent and added specification of k in l.48.}
+ \dubcheck{p.12, l.50: it is not clear to me what you mean by "constraining the fit to a linear trend", even after reading the following part; provide some additional explanations}
  + \response{We clarified what we meant by "constraining the fit to a linear trend" by adding explantion about participants being allowed to freely draw: "The 'You Draw It' method does not restrict participants to draw a straight line as other methods would, such as using a ruler. Instead, participants are allowed to freely draw a line with potential curvature. Using the `lmer` function in the lme4 package, a linear mixed model (LMM) was fit separately to the OLS residuals and PA residuals, emulating the effect of constraining participants to draw a linear trend."}

@@

  + \response{Selected one reference (book) which we used for the analysis.}
+ p.14, bottom - p.15, top: this model is very similar to the previous one; simply state y_i,j,k,drawn etc. are the same as for model (2) and only specify what is new/different here, e.g., the s_i
  + \response{Clarified similarities between the linear mixed model equation and the generalized additive mixed model equation by stating "Defining $e_{ijk,fit}$ the same as in equation (2) above, the GAMM equation...".}
+ \dubcheck{p.16, Discussion:} Add some references from cognitive/human movement sciences that human arm movement is a complex task. Drawing a straight line is not as easy as it sounds. Could this lead to the large residuals when x approaches 20 ?!? Here are some candidate articles (based on the abstracts). Do some further reading and find additional related references.

@@

  + https://doi.org/10.1007/BF00241501
  + https://doi.org/10.1145/2820619.2820633
  + \response{These are some really great suggestions that lead to some interesting and fun reads related to this work. We reviewed the suggested readings and followed some references to other related articles. A few sentences have been added to the discussion section to address some of the cognitive/human movement sciences and influcences on our study.}
+ p.17, Future work:

@@

  + This leads to a modified version of this study: Give the participants a horizontal line segment at the bottom of the graph and let them shift and rotate this line segment until they think they have found the best fit for the point cloud. This likely would remove some of the big errors we see in Figs 5 & 6 as x approaches 20.
  + A second option for future work that goes beyond Mosteller et al. (1981) is to create scatterplots with one (or multiple) extreme outliers. What would humans fit then? Still, the PCA line or rather a robust regression line that ignores the outliers (likely not the OLS line that may be totally off from a line that goes through most of the data). [Apparently, this will also depend on the instructions and the feedback that is provided in the training part of such a study.]
  + Neither of these has to be implemented in the current study, but listing both options as natural extensions for the straight line assessment seems to be worthwhile.
  + \response{These would be some great extensions to validating 'You Draw It' as a method for graphical testing. We added these suggestions as well as suggested a study which compares the indirect interaction (mouse and pointer on the screen) and direct interaction (digital pen on a tablet) method of using 'You Draw It'.}
+ \dubcheck{p.17-21, References:}

@@

  + Deming (1943) is incomplete (is this a book or a journal article?)
  + in Kosslyn & Kosslyn, spell out OUP (and indicate the city/state and not USA)
  + check capitalization of nouns; if I am not mistaken, JCGS uses upper case letters for all nouns [some of these apply to other references as well]
  + \response{Reviewed references and adjusted bibtex file and style.}

@@

In my opinion, the actual “novel” aspects of the presented research work are: 1) the fact that
an online tool initially meant for the general public can replicate empirical findings from the literature on graph perception; 2) the use of generalized additive mixed models. Concerning this second aspect, I am seriously wondering whether it was necessary at all to use such complex models to prove that participants minimized the orthogonal distance of the points to the fit. Simpler statistical analyses could have been (or could be) provided. However, since I am not an expert of these models, I recognize that it might be of interest for a specialized readership.

In conclusion, I strongly believe that the manuscript should concentrate on the actual novel aspects mentioned above (the use of an online tool and the use of generalized additive mixed models) and not on an aspect that the literature has already empirically tested and
statistically confirmed and formalized. If such a major revision is undertaken by the authors, the article might be worth publishing (especially within the scope of the journal).

Below I provide a series of other aspects that should also be considered.

+ Authors should avoid any form of plagiarism: for example, lines 43 to 46 (page 2) and line 55 (page 8) and line 22 (page 9) are identical excerpts from Ciccione & Dehaene.
  + \response{Thank you for catching this. Amongst literature review and edits, we did not realize we had directly quoted. We have rephrased the information in each of the indicated locations as they are important points to include.}
+ Page 3, line 20. The method of adjustment is introduced but not explained, leaving the reader wondering what it refers to.
  + \response{We describe the method of adjustment and added a reference.}
+ Page 3, line 50. The line-up protocol is introduced but, again, no clear explanation is provided.
  + \response{We added a couple sentences to describe the lineup protocol in addition to the Buja (2009) reference. "A statistical lineup is a plot consisting of smaller panels where the viewer is asked to identify the target panel containing the real data from among a set of decoy null plots which display data under the assumption there is no relationship. If the viewer can identify the target panel randomly embedded within the set of null panels, this suggests that the real data is visually distinct from data generated under the null model."}
+ Page 5, line 21. The concept of “ensemble perception” is not clearly related to what is described just before: how can the finding itself of human fitting a principal axis on scatterplot “support” (as the authors say) the work on ensemble perception? The link needs to be clarified or avoided.
  + \response{We removed the sentence discussing "ensemble perception" in relation to the Mosteller (1981) paper.}
+ Page 8, table 1. I think it’s a little bit confusing to see that the intercept parameter is called yx whereas in the equation in the text is called b0. The authors should choose one notation or
the other and stay consistent through the text.
  + \response{We aimed to replicate the trends from Mosteller et al. (1981). This original manuscript gives vague descriptions of the coefficients used to describe their trends. The way they defined their parameters was by using the y-value at the mean of the x-values $(y_{\bar{x}})$ and the slope. When simulated our data, we first generate and jitter x-values across the domain, therefore, x-bar may differ for each unique data set due to the random jitter. In turn, the actual y-intercept $(\beta_0)$ will vary slightly from data set to data set, but each trend will go through the same $(\bar{x}, y_{\bar{x}})$ point. We changed the equation notation to follow the point-slope equation rather than changing notation.}
+ Page 10, line 39. The reader is invited to refer to an R function for the description of the regression equation based on the principal component axis. Authors should provide an actual formalization of the equation.
  + \response{We reran the analysis using the mcreg function in the mcr package which implements Deming regression (equivalent to the principal axis regression). We clarified and provided a reference for this in the "Fitted Regressions" Results section.}
+ Page 12, line 48 and page 15, line 45. Again (as in page 5, line 21), the authors’ work is directly presented as providing support for ensemble perception. Ensemble perception simply states that human perception tends to average, in parallel, the information coming from various items/locations in the visual field. Evidence for a principal axis component extraction does not provide a direct piece of evidence for ensemble perception. Authors should better clarify this link or avoid introducing the concept of ensemble perception.
  + \response{We agree that ensemble perception is not necessarily the focus of this paper and have removed introducing the concept.}

@@

### Editor's Comments to Author

Thank you for your submission and interesting work. Two reviewers have provided feedback and suggestions on your manuscript. There are a few points I would like to emphasize and request clarification:

1. \dubcheck{In your manuscript, it would be helpful if you could clarify and emphasize the novelty of your experiment, data collection, and analysis. In particular, how does your work differ from Ciccione and Dehaene (2021)?(Ciccione, L. and Dehaene, S., 2021. Can humans perform mental regression on a graph? Accuracy and bias in the perception of scatterplots. Cognitive Psychology, 128, p.101406.)}
  + \response{We clarifed the differences between our work and Ciccione (2021) as our study use a more flexible method of testing graphics and introduces a way to analyze the data using GAMM's. While this study was designed to replicate results found in previous studies as a way to validate our method, our intent is to set the foundation for using 'You Draw It' for testing statistical graphics as the tool extends nicely to the non-linear setting. We added this clarification into the end of Section 1.2.}
2. \todo{In Section 3, it is helpful to note the R functions used as comments or footnotes, but you should also clearly describe the method of estimation. In other words, what are the R functions doing to estimate the model?}
  + \response{We added the method of estimations based on the R documentation for the lmer and bam functions.}
3. Could you clarify if the participants were known to be experts in statistics or related fields?
  + \response{We added a sentence describing that the data is from a pilot study and thus participants from the recruitment methods likely have an interest in advancing research in data visualization: "As this is a pilot study, participants from Twitter and Reddit pages related to data visualization voluntarily completed the study and likely have an interest in fields related to statistics and want to help advance research in graphics."}
4. Throughout the manuscript, references to “principle components” should be “principal components.”
  + \response{Based on a reviewer's suggestion, we have changed terminology to principal axis.}

### 2022-08-03 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
knitted manuscript revisions

```{r setup, include = F}
knitr::opts_chunk$set(
  echo = F, eval = T, messages = F, warnings = F,
  fig.width = 6, fig.height = 4,  fig.align = 'center',
  out.width = "\\linewidth", dpi = 300, 
  tidy = T, tidy.opts=list(width.cutoff=45),
  fig.pos = "tbp",
  out.extra = ""
)
```

@@

@@

In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing. Our visual system is naturally built to look for structure and identify patterns. For instance, points going down from left to right indicates a negative correlation between the $x$ and $y$ variables. Our research is intended to implement the 'You Draw It' feature as a way to measure the patterns we see in data.
We validate the 'You Draw It' method by replicating the less technological study conducted by @mosteller1981eye.
Based on previous research, we hypothesize that visual regression tends to mimic regression based on the principal axis rather than an ordinary least squares regression.
In order to assess this hypothesis, we introduce a method for statistically modeling the participant drawn lines using generalized additive mixed models.

# Methods

@@

Participants were recruited through through Twitter, Reddit, and direct email in May 2021.
A total of 35 individuals completed 119 unique 'You Draw It' task plots; data collected from all completed 'You Draw It' task plots were included in the analysis.
All participants had normal or corrected to normal vision and signed an informed consent form.
The experimental tasks took approximately 15 minutes to complete.
While this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
These data were collected to validate this method of graphical testing, with the hopes of providing a new tool to assess graphical perception interactively.
Participants completed the experiment on their own computers in an environment of their choosing. The experiment was conducted and distributed through a Shiny application [@shinyPkg] found at [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).

@@

eyefitting_sim_data <- read_csv("analysis/eyefitting-simulated-data.csv")

@@

# Results

@@

x1tox2 = coef(lm(x1~x2,dat))
x2tox1 = coef(lm(x2~x1,dat))
slopeData = data.frame(slope = c(x1x2cor,x2tox1[2]),
                       type=c("PC", "OLS"))

@@

segmeData = rbind(data.frame(x=dat[,1],y=dat[,2],xend=points[1,],yend=points[2,],type = "PC"),
                  data.frame(x=dat[,1],y=dat[,2],yend=dat[,1]*x2tox1[2],xend=dat[,1],type="OLS"))

@@

```{r eyefitting-plots}
eyefitting_simulated_data <- read_csv("analysis/eyefitting-simulated-data.csv")

eyefitting_model_data <- read.csv("analysis/eyefitting-model-data.csv")

@@

```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Illustrates the data associated with and collected for one `You Draw It' task plot. Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PA regression line based on the principal axis (solid orange).", out.width="50%"}
# trial.feedback <- read.csv("results/youdrawit-eyefitting-example-feedback.csv") %>%
#     mutate(`Parameter Choice` = "F")
# trial.sim <- read.csv("results/youdrawit-eyefitting-example-simulated.csv") %>%
#     mutate(`Parameter Choice` = "F")
#     
# trial.feedback %>%
#   ggplot(aes(x = x)) +
#   geom_point(data = trial.sim, aes(y = y), alpha = 0.7) +
#   geom_line(aes(y = y, color = "OLS", linetype = "OLS")) +
#   geom_line(aes(y = ypca, color = "PA", linetype = "PA")) +
#   geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
#   facet_wrap(~`Parameter Choice`) +
#   theme_bw(base_size = 14) +
#   theme(aspect.ratio = 1,
#         legend.position = "bottom",
#         axis.text    = element_text(size = 12),
#         axis.title   = element_text(size = 12),
#         legend.title = element_text(size = 12),
#         legend.text  = element_text(size = 12),
#         # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
#         # strip.background = element_rect(size = 0.8),
#         legend.key.size = unit(1, "line")
#         ) +
#   scale_x_continuous(limits = c(0,20)) +
#   scale_y_continuous(limits = c(-5, 17), breaks = seq(-5, 15, 5)) +
#   scale_color_manual("", values = c("black", "steelblue", "orange")) +
#   scale_linetype_manual("", values = c("dashed", "solid", "solid"))

eyefitting_model_data %>%
  filter(participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F") %>%
  rename(`Parameter Choice` = parm_id) %>%
  ggplot(aes(x = x)) +
  geom_line(aes(y = yols, color = "OLS", linetype = "OLS")) +
  geom_line(aes(y = ypca, color = "PA", linetype = "PA")) +
  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
  geom_point(data = eyefitting_simulated_data %>%
               filter(dataset == "point_data", participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F")  %>% rename(`Parameter Choice` = parm_id),
             aes(x = x, y = y)) +
  facet_wrap(~ `Parameter Choice`, labeller = label_both) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  scale_y_continuous("y") +
  scale_color_manual("", values = c("black", "steelblue", "orange2")) +
  scale_linetype_manual("", values = c("dashed", "solid", "solid"))
```

### Linear Trend Constraint

Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) was fit separately to the OLS residuals and PA residuals, constraining the fit to a linear trend. 
Parameter choice, $x$, and the interaction between $x$ and the parameter choice were treated as fixed effects with a random participant effect included to account for variation due to participant.
The LMM equation for each fit (OLS and PA) is given by:
\begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation}
\noindent where

@@

Eliminating the linear trend constraint, the `bam` function in the mgcv package [@mgcv_pkg] was used to fit a generalized additive mixed model (GAMM) separately to the OLS residuals and PA residuals to allow for estimation of smoothing splines.
Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice. A random participant effect was included to account for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
Defining e_{ijk,fit} the same as in equation (2) above, the GAMM equation for each fit (OLS and PA) residuals is given by:
\begin{equation}
e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
\end{equation}
\noindent where

@@

# Discussion and Conclusion

The intent of this research was to adapt 'You Draw It' from the New York Times feature as a tool and method for testing graphics and introduce a method for statistically modeling the participant drawn lines.
We provided support for the validity of the 'You Draw It' method by replicating the study found in @mosteller1981eye.
Using generalized additive mixed models, we assessed the deviation of the participant drawn lines from the statistically fitted regression lines.
Our results found that when shown points following a linear trend, participants visually fit a regression line that mimics the first principal axis regression as opposed to ordinary least squares regression. 
Data simulated with a larger variance provided strong support for a participants tendency to visually fit the first principal axis regression.
We utilized modern technology to replicate a study conducted 40 years ago, and strengthened the original results with current analysis methods which allow for more flexibility and sophistication.
Our results indicate that participants minimized the distance from their drawn regression line over both the $x$ and $y$ axis simultaneously. 
<!-- These results provide support that humans perform "ensemble perception" in a statistical graphic setting. -->
We allowed participants to draw trend lines that deviated from a straight line and gained an insight into the curvature the human eye perceives in a set of points.

# Future Work

This study provided a basis for the use of 'You Draw It' as a tool for testing statistical graphics and introduced a method for statistically modeling participant drawn lines using generalized additive mixed models.
Further investigation is necessary to implement this method in non-linear settings and with real data in order to facilitate scientific communication.
This tool could also be used to evaluate human ability to extrapolate data from trends. 
In the future, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.

@@

+ **Participant Data:** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data). 
+ **Data Analysis Code:** The code used to replicate the analysis in this paper can be found at [earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
+ **Study Applet:** The shiny app used to conduct the study can be accessed at  [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
+ **RShiny Applet Code:** The code used to create the RShiny Applet for data collection can be found at [github.com/earobinson95/you-draw-it-validation-applet](https://github.com/earobinson95/you-draw-it-validation-applet).

>>>>

```{r setup, include = F}
knitr::opts_chunk$set(
  echo = F, eval = T, message = F, warning = F,
  fig.width = 6, fig.height = 4,  fig.align = 'center',
  out.width = "\\linewidth", dpi = 300, 
  tidy = T, tidy.opts=list(width.cutoff=45),
  fig.pos = "tbp",
  out.extra = ""
)
```

@@

```{r eyefitting-data}
eyefitting_simulated_data <- read_csv("analysis/eyefitting-simulated-data.csv")

eyefitting_model_data <- read.csv("data/eyefitting-model-data.csv")

@@

In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a new tool for graphical testing. 
Our visual system is naturally built to look for structure and identify patterns. 
For instance, points going down from left to right indicates a negative correlation between the $x$ and $y$ variables. 
Our research is intended to implement the 'You Draw It' feature as a way to measure the patterns we see in data.
The graphical testing method used in this study differs from prior methods found in @mosteller1981eye and @ciccione2021can by allowing participants to freely draw estimated trend lines - a method which extends nicely to a nonlinear setting.
We validate the 'You Draw It' method by replicating the less technological study conducted by @mosteller1981eye.
In [Section 2](#methods) we describe our participant sample, the graphical task to be completed, and the data generation process and study design.
[Section 3](#results) describes the participant data collected and shares results from the analyses of the data using mixed models.
Overall conclusions and discussion of results are presented in [Section 4](#conclusion-discussion) with extensions to the current work suggested in [Section 5](#future-work). 
The RShiny applet to complete the study, participant data used for analysis, and code to replicate the analysis can be found in the [Supplementary Material](#supplementary-material).
Based on previous research, we hypothesize that visual regression tends to mimic regression based on the principal axis rather than an ordinary least squares regression.
In order to assess this hypothesis, we introduce a method for statistically modeling the participant drawn lines using generalized additive mixed models (GAMM).
While the focus of this paper is to share the results from the validation study which uses the new 'You Draw It' method to evaluate visually fitted linear trends to statistical regression results, the intent of this work is to set the foundation and demonstrate the strength of the combination of the 'You Draw It' method and GAMM analyses for testing statistical graphics and to extend the use of the method beyond the linear setting.

# Methods {#methods}

@@

Participants were recruited through through Twitter, Reddit, and direct email in May 2021.
A total of 35 individuals completed `r length(unique(eyefitting_model_data$plot_id))` unique 'You Draw It' task plots.
Data were collected as a part of a pilot study meant to test the applet; therefore, either voluntary participant dropout or disconnection from a server not designed to accommodate large magnitudes of participants resulted in missing plots in our data set for analysis.
All participants had normal or corrected to normal vision and signed an informed consent form.
The experimental tasks took approximately 15 minutes to complete.
As this is a pilot study, participants from Twitter and Reddit pages related to data visualization voluntarily completed the study and likely have an interest in fields related to statistics and want to help advance research in graphics.
While this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
These data were collected to validate this method of graphical testing, with the hopes of providing a new tool to assess graphical perception interactively.
Participants completed the experiment on their own computers in an environment of their choosing. The experiment was conducted and distributed through a Shiny application [@shinyPkg] found at [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).

@@

eyefitting_sim_data <- read_csv("analysis/eyefitting-simulated-data-example.csv")

@@

# Results {#results}

@@

x1tox2 = coef(lm(x1~x2,dat))
x2tox1 = coef(lm(x2~x1,dat))
slopeData = data.frame(slope = c(x1x2cor,x2tox1[2]),
                       type=c("PA", "OLS"))

@@

segmeData = rbind(data.frame(x=dat[,1],y=dat[,2],xend=points[1,],yend=points[2,],type = "PA"),
                  data.frame(x=dat[,1],y=dat[,2],yend=dat[,1]*x2tox1[2],xend=dat[,1],type="OLS"))

@@

@@

```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Illustrates the data associated with and collected for one `You Draw It' task plot. Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PA regression line based on the principal axis (solid orange).", out.width="50%"}
eyefitting_model_data %>%
  filter(participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F") %>%
  dplyr::rename(`Parameter Choice` = parm_id) %>%
  ggplot(aes(x = x)) +
  geom_line(aes(y = yols, color = "OLS", linetype = "OLS")) +
  geom_line(aes(y = ypca, color = "PA", linetype = "PA")) +
  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
  geom_point(data = eyefitting_simulated_data %>%
               filter(dataset == "point_data", participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F") %>% dplyr::rename(`Parameter Choice` = parm_id),
             aes(x = x, y = y)) +
  facet_wrap(~ `Parameter Choice`, labeller = label_both) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  scale_y_continuous("y") +
  scale_color_manual("", values = c("black", "steelblue", "orange2")) +
  scale_linetype_manual("", values = c("dashed", "solid", "solid"))
```

### Linear Trend Constraint
<!-- https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf -->

The 'You Draw It' method does not restrict participants to draw a straight line as other methods would, such as using a ruler. Instead, participants are allowed to freely draw a line with potential curvature.
Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) was fit separately to the OLS residuals and PA residuals, emulating the effect of constraining participants to draw a linear trend.
Both fixed and random parameter estimates in the LMM were determined by optimizing the restricted maximum likelihood (REML) through penalized least squares.
Parameter choice, $x$, and the interaction between $x$ and the parameter choice were treated as fixed effects with a random participant effect included to account for variation due to participant.
The LMM equation for each fit (OLS and PA) is given by:
\begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation}
\noindent where

@@

Eliminating the linear trend constraint, the `bam` function in the mgcv package [@mgcv_pkg] was used to fit a generalized additive mixed model (GAMM) separately to the OLS residuals and PA residuals to allow for estimation of smoothing splines.
The `bam` function is used to fit GAMM's to very large data sets and use lower memory than the `gam` function.
Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice. A random participant effect was included to account for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
Defining e_{ijk,fit} the same as in equation (2) above, the GAMM equation for each fit (OLS and PA) residuals is given by:
\begin{equation}
e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
\end{equation}
\noindent where

@@

# Discussion and Conclusion {#conclusion-discussion}

The intent of this research was to adapt 'You Draw It' from the New York Times feature as a tool and method for testing graphics and introduce a method for statistically modeling the participant drawn lines.
We provided support for the validity of the 'You Draw It' method by replicating the study found in @mosteller1981eye.
Using generalized additive mixed models, we assessed the deviation of the participant drawn lines from the statistically fitted regression lines.
Our results found that when shown points following a linear trend, participants visually fit a regression line that mimics the first principal axis regression as opposed to ordinary least squares regression. 
Data simulated with a larger variance provided strong support for a participants tendency to visually fit the first principal axis regression.
We utilized modern technology to replicate a study conducted 40 years ago, and strengthened the original results with current analysis methods which allow for more flexibility and sophistication.
Our results indicate that participants minimized the distance from their drawn regression line over both the $x$ and $y$ axis simultaneously. 
<!-- These results provide support that humans perform "ensemble perception" in a statistical graphic setting. -->
We allowed participants to draw trend lines that deviated from a straight line and gained an insight into the curvature the human eye perceives in a set of points.
Researchers in cognitive and human movement sciences have found that human arm movement is a complex task [@miall1995curvature; @rousset2015study].
The 'You Draw It' method described in this paper uses *indirect interaction* in which the mouse position and resulting visual line on the screen are dissociated.
Therefore, curvature found in participant drawn lines from a straight lines could potentially be explained by the lack of coordination which results from the eye-hand dissociation from indirect drawing and the distortion of visual perception affecting the curvature of movements.
Additionally, there is a training effect related to the completion of the 'You Draw It' task - the movement of the line must be slow so that the visual representation on the screen can accurately capture each movement. 
@de1991misdirections conducted a study in which participants moved their hand slowly from an initial position in front of them to a visual target (movement task); they were then asked to repeat the task using different sizes of pointers (perceptual task).
Their results indicated that deviations from the shortest pointers were comparable to those of the movement task, but that bias increased as the length of the pointer increased.
While we suggested participants use a mouse to complete the study, we could not require the use; therefore, some participants may have used a track-pad and results may have been influenced by the pressure placed on their track-pad [@easton1978finger].

# Future Work {#future-work}

This study provided a basis for the use of 'You Draw It' as a tool for testing statistical graphics and introduced a method for statistically modeling participant drawn lines using generalized additive mixed models.
Additional studies related to the validation and use of the tool would be useful for providing insight into explanations of biases introduced by the task such as the deviation from a straight line.
For instance, a variation on the current study could compare manual adjustment methods such as shifting and rotating a horizontal line segment until the fit is suitable to the 'You Draw It' method on the same set of data. This might explain the large deviation from the participant drawn line as $x$ approaches 20.
Another useful extension study would be to compare the 'You Draw It' method as conducted by direct interaction - using a digital pen on a tablet - to indirect interaction - using a computer mouse to relate to a pointer on the screen. 
Further extensions to this work might ask participants to draw a trend-line through scatterplots with one (or multiple) extreme outliers in order to evaluate the perceptual system's resistance to outliers.

While the focus of this study was on drawing linear trend-lines, further investigation is necessary to implement this method in non-linear settings and with real data in order to facilitate scientific communication - a strength of the combination of the flexible 'You Draw It' method and GAMM analysis method.
This tool could also be used to evaluate human ability to extrapolate data from trends. 
In the future, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.

@@

{#supplementary-material}

+ **Study Applet:** The shiny app used to conduct the study can be accessed at  [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
+ **RShiny Applet Code:** The code used to create the RShiny Applet for data collection can be found at [github.com/earobinson95/you-draw-it-validation-applet](https://github.com/earobinson95/you-draw-it-validation-applet).
+ **Participant Data:** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data). 
+ **Data Analysis Code:** The code used to replicate the analysis in this paper can be found at [earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
